\documentclass[10pt,letterpaper,twoside]{book}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage[margin=1.7cm,bottom=2.2cm,headsep=10pt]{geometry}
\usepackage{mathpazo}
\usepackage{amsmath,amssymb}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{xcolor}
\usepackage[most]{tcolorbox}
\usepackage{graphicx}
\usepackage{fancyhdr}
\usepackage{enumitem}
\usepackage[hidelinks,pdfusetitle]{hyperref}
\usepackage{titlesec}

% --- angle macros used throughout the generated tables ---
\newcommand{\dg}{\ensuremath{{}^{\circ}}}
\newcommand{\pr}{\ensuremath{{}^{\prime}}}

\setlength{\tabcolsep}{4pt}
\renewcommand{\arraystretch}{1.05}
\emergencystretch=4em
\hyphenpenalty=1000

\pagestyle{fancy}
\fancyhf{}
\fancyhead[LE,RO]{\small\thepage}
\fancyhead[RE]{\small\itshape The Sextant Navigator's Almanac 2026}
\fancyhead[LO]{\small\itshape\leftmark}
\renewcommand{\headrulewidth}{0.3pt}

\titleformat{\chapter}[display]{\normalfont\huge\bfseries}{\chaptertitlename\ \thechapter}{16pt}{\Huge}

% inline data macros produced by generate_tables.wls
\input{tables/macros.tex}

\title{The Sextant Navigator's Almanac \& Sight-Reduction Handbook (2026)}
\author{Generated from the open-source \textsc{SextantNavigation} ephemeris engine}
\date{Edition of \genDate}

\begin{document}
\frontmatter

% ============================ TITLE PAGE ============================
\begin{titlepage}
\centering
\vspace*{2.2cm}
{\Huge\bfseries The Sextant Navigator's\\[4pt] Almanac 2026\par}
\vspace{6pt}
{\Large\itshape \& Sight-Reduction Handbook\par}
\vspace{1.4cm}
{\large A self-contained nautical almanac and paper sight-reduction manual,\\
computed entirely from a validated, open-source celestial-navigation engine.\par}
\vfill
{\large Greenwich Hour Angle and Declination of the Sun,\\
GHA of Aries, and the 58 navigational stars, for the year 2026 (UT).\par}
\vspace{1.2cm}
{\normalsize Generated from \textsc{SextantNavigation}\\
\url{https://github.com/mthiel74/SextantNavigation}\par}
\vspace{0.6cm}
{\small Edition of \genDate\par}
\vspace{0.4cm}
{\footnotesize \textcopyright{} 2026 Marco Thiel. Licensed under Creative Commons
Attribution 4.0 International (CC BY 4.0).\par}
\vfill
{\footnotesize\bfseries FOR EDUCATION AND PRACTICE ONLY --- NOT FOR SOLE-MEANS NAVIGATION\par}
\end{titlepage}

% ============================ DISCLAIMER ============================
\thispagestyle{empty}
\vspace*{1.5cm}
\begin{tcolorbox}[colback=red!5!white,colframe=red!60!black,boxrule=1.2pt,
  arc=3pt,title=\bfseries IMPORTANT --- READ BEFORE USE,fonttitle=\large]
\large
This almanac and handbook are an \textbf{educational deliverable}. Every number
in it was computed offline from the open-source ephemeris engine at
\url{https://github.com/mthiel74/SextantNavigation}, not copied from an official
publication.

\medskip
\textbf{Accuracy.} The underlying engine agrees with JPL for the Sun's geographic
position to about \textbf{0.14$\pr$}, reproduces Meeus' lunar worked example to
0.002$\pr$ (geocentric), and precesses the star catalogue with IAU--1976
precession plus proper motion. Realistic end-to-end accuracy for a reduced sight
is on the order of \textbf{one arcminute (about one nautical mile)}. That is good
enough to \emph{learn and practise} celestial navigation, and to check your paper
arithmetic --- but it is \textbf{not} certified for life-safety use.

\medskip
\textbf{Do not use this document as your sole means of navigation at sea.} For
real voyages, use an official \emph{Nautical Almanac} (HM Nautical Almanac Office
/ US Naval Observatory) and cross-check every figure. The authors accept no
liability for any use of this material.
\end{tcolorbox}

\bigskip
\noindent{\small\textbf{Sources \& method.} The positions in this book are
\emph{computed with the Wolfram Language}, not reproduced from any published
almanac. The Sun's geographic
position is from the low-precision solar formulae of the \emph{Astronomical
Almanac}; the Moon from Meeus's abridged ELP theory (\emph{Astronomical
Algorithms}, ch.~47); Greenwich sidereal time from the IAU GMST formula; and the
star places are propagated from \emph{Hipparcos} catalogue (ESA) J2000 mean
positions using IAU--1976 precession and catalogue proper motions to epoch~2026.5.
Full source code and tests: \url{https://github.com/mthiel74/SextantNavigation}.}

\medskip
\noindent{\small\textbf{Acknowledgements.} All computations and tables were
produced with the \emph{Wolfram Language} (Wolfram Research). This almanac
gratefully builds on published astronomy: the star catalogue derives from the
ESA \emph{Hipparcos} mission; the lunar theory follows Jean Meeus,
\emph{Astronomical Algorithms}; and the solar and sidereal-time formulae follow
the conventions of the IAU and the \emph{Astronomical Almanac} (HM Nautical
Almanac Office / US Naval Observatory). It is an independent, open-source
companion to the article cited above, and is not affiliated with, or endorsed
by, any of these organisations.}

\vfill
\begin{center}\small
Source engine, tests, and this document: \url{https://github.com/mthiel74/SextantNavigation}\\
Companion article: \url{https://community.wolfram.com/groups/-/m/t/3743363}
\end{center}
\clearpage

% ============================ HOW TO USE ============================
\chapter*{How to use this almanac}
\addcontentsline{toc}{chapter}{How to use this almanac}
\markboth{How to use this almanac}{How to use this almanac}

All times in the data tables are \textbf{Universal Time (UT)}, the time kept by
your chronometer set to Greenwich. All angles are given in degrees and decimal
minutes to $0.1\pr$, the same precision as a commercial nautical almanac.

\paragraph{To reduce a sight you need four numbers for the instant of your
observation:} the body's \emph{Greenwich Hour Angle} (GHA) and \emph{Declination}
(Dec), your \emph{assumed position}, and your \emph{observed altitude} $H_o$.
This book supplies the first two:

\begin{itemize}[nosep]
\item \textbf{Sun.} Turn to Part~II, find the month, then the row for your day and
      the whole hour of UT. Read GHA and Dec directly. For the odd minutes and
      seconds, add the increment from Part~IV (the Sun's GHA grows a steady
      $15\dg$ per hour). Declination changes so slowly you may interpolate it by eye.
\item \textbf{A star.} Turn to Part~III. Read \textbf{GHA Aries} for your day and
      hour, then add the star's \textbf{SHA} (Sidereal Hour Angle) from the star
      table: $\text{GHA}_\star = \text{GHA Aries} + \text{SHA}$. The star's
      declination is taken straight from the table. Add the Part~IV increment for
      minutes and seconds (Aries runs at $15.041\dg$ per hour).
\end{itemize}

\paragraph{Then work the triangle.} Part~I explains, step by step and with fully
worked 2026 examples, how to correct your sextant reading and turn these numbers
into a line of position on your chart.

\tableofcontents

\mainmatter

% ============================================================
% PART I
% ============================================================
\part{Finding Yourself with a Sextant}
\chapter{The idea: circles of equal altitude}

When you measure the altitude of a celestial body --- the angle between it and
the horizon --- you learn one thing about where you are: your distance from the
body's \emph{geographic position} (GP), the point on the Earth's surface directly
beneath it. If the Sun's measured altitude is $H_o$, then your \emph{zenith
distance} is
\[
  z = 90\dg - H_o ,
\]
and $z$, expressed in arcminutes, is exactly your great-circle distance from the
GP in nautical miles (one minute of arc on the Earth $=$ one nautical mile). You
must therefore lie somewhere on a circle of radius $z$ centred on the GP: a
\textbf{circle of equal altitude}.

One sight gives a circle. A \emph{second} sight of another body (or the same body
later) gives a second circle, and two circles cross at two points. A \emph{third}
sight resolves which of the two intersections is really you. In practice we never
draw these enormous circles; near your estimated position each circle is
indistinguishable from its tangent straight line, the \textbf{line of position}
(LOP), and we plot those instead. That is the method of this chapter.

\section{What you need}
\begin{itemize}[nosep]
\item A \textbf{sextant} to measure altitude $H_s$ above the sea horizon.
\item An accurate \textbf{clock on UT} (a chronometer). An error of 4 seconds of
      time moves your fix about one nautical mile in longitude.
\item This \textbf{almanac}, for GHA and Dec of the body.
\item A \textbf{dead-reckoning (DR) position}: your best estimate of where you are,
      used to pick an ``assumed position'' close by.
\item Plotting sheet, dividers, parallel rule, and a pocket calculator with
      trigonometric functions (or the concise sight-reduction tables).
\end{itemize}

\chapter{Correcting the sight}

The raw sextant altitude $H_s$ is not yet the true altitude of the body's centre.
Apply, in order:

\begin{enumerate}[nosep]
\item \textbf{Index error} $I$. Your own sextant's zero error, found by sighting
      the horizon. Subtract if ``on the arc'', add if ``off the arc''.
\item \textbf{Dip} of the horizon. Because your eye is above the sea, the visible
      horizon is depressed below the true horizontal by
      \[ \text{Dip} = 1.76\sqrt{h}\ \text{arcminutes}, \qquad h=\text{height of eye in metres.} \]
      Always subtract. (See the table in Part~IV.)
\item \textbf{Refraction} $R$. The atmosphere bends the light down, raising the
      apparent altitude; subtract $R$. This almanac uses Bennett's formula,
      $R = \cot\!\big(H_a + 7.31/(H_a+4.4)\big)$ arcminutes, evaluated at the
      apparent altitude $H_a$ (altitude after index error and dip). Refraction is
      about $1\pr$ at $45\dg$ but $10\pr$ near the horizon.
\item \textbf{Semidiameter} (Sun and Moon). You bring the \emph{limb} (edge), not
      the centre, to the horizon. Add the semidiameter (SD) for a lower-limb sight,
      subtract it for an upper-limb sight. The Sun's SD is about $16\pr$; the exact
      monthly value is tabulated in Part~IV.
\item \textbf{Parallax} (Moon, and slightly the Sun). A near body is seen against
      the stars from a slightly different direction than it would be from the
      Earth's centre. The correction, always \emph{added}, is
      \[ P = \arcsin\!\big(\sin(\mathrm{HP})\cos H\big) \approx \mathrm{HP}\cos H, \]
      where HP is the horizontal parallax. For the Moon HP is about $57\pr$, so the
      parallax reaches nearly a full degree; for the Sun it is a negligible
      $0.1\pr$. Stars are effectively infinitely far away: no parallax, no
      semidiameter.
\end{enumerate}

The result is the \textbf{observed altitude}
\[
  H_o = H_s - I - \text{Dip} - R \;(\pm\,\text{SD})\;(+\,P).
\]

\chapter{Sight reduction: the Marcq St-Hilaire intercept}

We cannot solve for position directly, so we \emph{guess} an assumed position (AP)
close to our DR and ask: \emph{if I were at the AP, what altitude and bearing
would the body have?} Comparing that computed altitude $H_c$ with the observed
$H_o$ tells us how far, and in which direction, to shift from the AP onto our true
LOP.

\section{The navigational (PZX) triangle}
On the celestial sphere, three points form a spherical triangle: the elevated
pole $P$, your zenith $Z$, and the body $X$. Its sides are the co-latitude
$90\dg-\varphi$, the co-declination $90\dg-\delta$, and the co-altitude
$90\dg-H_c$; the angle at the pole is the \textbf{local hour angle}
$\text{LHA}=\text{GHA}-\text{(west longitude)}$ (equivalently $\text{GHA}+$ east
longitude). Applying the spherical law of cosines to the side $90\dg-H_c$:
\begin{equation}
  \boxed{\;\sin H_c = \sin\varphi\,\sin\delta + \cos\varphi\,\cos\delta\,\cos(\text{LHA})\;}
\end{equation}
where $\varphi$ is the assumed latitude and $\delta$ the declination. The body's
\textbf{azimuth} $Z_n$ (true bearing, measured clockwise from north) comes from
\begin{equation}
  \cos Z = \frac{\sin\delta - \sin H_c\,\sin\varphi}{\cos H_c\,\cos\varphi},
  \qquad
  Z_n = \begin{cases} Z & \text{if } \sin(\text{LHA})<0 \ (\text{LHA }180\dg\text{--}360\dg),\\
                      360\dg - Z & \text{if } \sin(\text{LHA})>0 \ (\text{LHA }0\dg\text{--}180\dg).\end{cases}
\end{equation}

\section{The intercept, and how to plot it}
The heart of the method is a single subtraction. Because one minute of altitude
equals one nautical mile of distance from the GP,
\begin{equation}
  \boxed{\; a = 60\,(H_o - H_c)\ \text{nautical miles}\;}
\end{equation}
(with $H_o,H_c$ in degrees). If $H_o>H_c$ you are \emph{closer} to the body than
the AP is, so step the intercept \textbf{Toward} the body along the azimuth $Z_n$;
if $H_o<H_c$, step it \textbf{Away} (along $Z_n+180\dg$). A useful mnemonic is
\emph{``Ho Mo To''} --- $H_o$ More means Toward.

\textbf{To plot:} mark the AP on your sheet. Draw the azimuth line through it in
the direction $Z_n$. Measure $a$ nautical miles along that line, Toward or Away.
Through that point draw the LOP \emph{perpendicular} to the azimuth --- that
perpendicular is your line of position. Repeat for a second body; where the two
LOPs cross is your fix. A third LOP turns the small ``cocked hat'' triangle into a
confident position.

\section{Choosing the assumed position}
To keep the arithmetic clean, navigators choose the AP cleverly: take the
\emph{assumed latitude} as the whole degree of latitude nearest your DR, and the
\emph{assumed longitude} with whatever minutes make the LHA come out to a whole
number of degrees. Both examples below use this trick.

\chapter{Worked example 1: a Sun sight}
\label{ch:ex1}

At sea near $40\dg$N, $30\dg$W on 2026 June~29, you take a lower-limb Sun sight.
Your chronometer reads exactly \textbf{15:00:00 UT}. Height of eye is $3.0$ m and
your sextant's index error is $2.0\pr$ on the arc. Work as follows (all almanac
figures are read from Parts~II and~IV of this book).

\begin{center}\input{tables/example1_work.tex}\end{center}

\noindent\textbf{Reading the result.} The observed altitude exceeds the computed
altitude by $H_o-H_c=\exOneInt\pr$, so the intercept is \textbf{\exOneInt\ nautical
miles \exOneDir} the Sun. Plot the assumed position ($40\dg$N, $30\dg06.9\pr$W),
draw the azimuth line toward $Z_n=\exOneZn$, step \exOneInt~nm along it \exOneDir\
the Sun, and draw your line of position at right angles to that azimuth. A single
Sun sight gives one LOP; a second sight a few hours later (when the Sun's azimuth
has swung round) crosses it to give a running fix.

\emph{Check it yourself:} $\sin H_c = \sin40\dg\sin(23\dg12.0\pr) +
\cos40\dg\cos(23\dg12.0\pr)\cos(14\dg)$ gives $H_c=\exOneHc$, exactly the value in
the workform. The whole-degree LHA of $\exOneLHA$ is what the assumed-longitude
choice was designed to produce.

\chapter{Worked example 2: latitude by the noon Sun}

The simplest sight of all needs no azimuth and almost no arithmetic. When the Sun
crosses your meridian (\emph{local apparent noon}, LAN) it bears due north or due
south and reaches its greatest altitude. At that instant your latitude follows
from the meridian altitude alone.

At $30\dg$W the Sun transits your meridian when its GHA equals $30\dg$; from
Part~II that occurs on 2026 June~29 at the UT shown below. Measure the Sun's
maximum altitude, correct it to $H_o$ as before, and take the zenith distance
$z=90\dg-H_o$. Because the Sun is south of you (its declination $23\dg$N is less
than your latitude), your latitude is
\[
  \varphi = \delta + z \qquad (\text{body to the south}).
\]

\begin{center}\input{tables/example2_work.tex}\end{center}

\noindent The result, $\varphi=\exTwoLat$, agrees with the true position to a
tenth of a minute. (Had the Sun been \emph{north} of you, the rule would be
$\varphi=\delta-z$; if you straddle the sub-solar latitude,
$\varphi=z-\delta$ with the Sun to the south of an overhead-passing Sun. Always
sketch the meridian to get the sign right.)

\chapter{Worked example 3: a star sight}

Stars are reduced exactly like the Sun, but their GHA is built in two pieces. On
2026 January~15 at \textbf{22:00:00 UT}, from an assumed position $30\dg$N, you
shoot the star \textbf{Sirius}. Read GHA Aries from Part~III and Sirius's SHA and
Dec from the star table, then add:

\begin{center}\input{tables/example3_work.tex}\end{center}

\noindent Because a star has no semidiameter and no parallax, its sight correction
is only index error, dip, and refraction --- simpler than the Sun. The computed
altitude $H_c=\exThreeHc$ and azimuth $Z_n=\exThreeZn$ are then compared with your
observed $H_o$ exactly as in Example~1 to get the intercept.

\chapter{A note on longitude and time}

Latitude falls out of a noon sight almost for free, but \textbf{longitude is a
question of time}. The Earth turns $15\dg$ of longitude per hour, so every second
of chronometer error displaces your computed longitude by a quarter of a minute of
arc ($\approx 0.25\pr\cos\varphi$ nautical miles). This is why the marine
chronometer was one of history's great navigational prizes: to find longitude you
must know GMT precisely and compare it with your \emph{local} time as read from the
Sun or stars. In the intercept method above, the chronometer time enters through
the GHA you look up; an error in time is indistinguishable from an error in
longitude. Keep your clock honest, log its rate, and never reduce a sight without
the exact UT of the observation.

% ============================================================
% PART II  --- THE SUN
% ============================================================
\part{The Sun, 2026}
\chapter{Daily data: meridian passage, equation of time, semidiameter}

For each day of 2026 the table below gives the \textbf{meridian passage} of the
Sun at Greenwich (the UT at which the Sun's GHA is $0\dg$, i.e.\ it crosses the
Greenwich meridian), the \textbf{equation of time} (EoT $= 12^{\text{h}} -$
Mer.Pass.\ $=$ apparent $-$ mean solar time; a positive value means the sundial is
ahead of the clock), and the Sun's \textbf{semidiameter}. Note the semidiameter is
largest in early January (Earth at perihelion) and smallest in early July.

{\setlength{\tabcolsep}{3pt}\input{tables/sun_daily.tex}}

\chapter{Hourly GHA and Declination of the Sun}

Read the Greenwich Hour Angle and Declination of the Sun's centre for the day and
whole hour of UT; interpolate the minutes and seconds with the Sun increment table
in Part~IV. Entries run left to right, then top to bottom, in three columns of
day~/~hour~/~GHA~/~Dec.

\input{tables/sun_01.tex}
\input{tables/sun_02.tex}
\input{tables/sun_03.tex}
\input{tables/sun_04.tex}
\input{tables/sun_05.tex}
\input{tables/sun_06.tex}
\input{tables/sun_07.tex}
\input{tables/sun_08.tex}
\input{tables/sun_09.tex}
\input{tables/sun_10.tex}
\input{tables/sun_11.tex}
\input{tables/sun_12.tex}

% ============================================================
% PART III --- ARIES & STARS
% ============================================================
\part{GHA Aries \& the Navigational Stars, 2026}
\chapter{The 58 navigational stars}

For each star the table gives the \textbf{Sidereal Hour Angle} (SHA, the star's
angular distance west of the First Point of Aries) and \textbf{Declination} at
epoch 2026.5, together with the visual magnitude. These are \emph{mean places} for
epoch 2026.5; precession and proper motion shift them by up to about half an
arcminute toward the ends of the year --- within this almanac's $\sim\!1\pr$ budget,
but re-generate the tables from the source engine if you need another year. To find a star's
GHA at your moment of observation, add its SHA to the GHA of Aries from the next
chapter:
\[
  \text{GHA}_\star = \text{GHA Aries} + \text{SHA}\quad(\text{reduce modulo }360\dg).
\]

\input{tables/stars.tex}

\chapter{Hourly GHA of Aries}

The Greenwich Hour Angle of the First Point of Aries ($\gamma$) for each day and
whole hour of UT. Add the star's SHA (previous chapter) for its GHA, and the Aries
increment from Part~IV for the odd minutes and seconds. Entries run left to right,
then top to bottom, in four columns of day~/~hour~/~GHA~$\gamma$.

\input{tables/aries_01.tex}
\input{tables/aries_02.tex}
\input{tables/aries_03.tex}
\input{tables/aries_04.tex}
\input{tables/aries_05.tex}
\input{tables/aries_06.tex}
\input{tables/aries_07.tex}
\input{tables/aries_08.tex}
\input{tables/aries_09.tex}
\input{tables/aries_10.tex}
\input{tables/aries_11.tex}
\input{tables/aries_12.tex}

% ============================================================
% PART IV --- CORRECTIONS & INCREMENTS
% ============================================================
\part{Correction \& Increment Tables}
\chapter{Altitude corrections}

\section{Dip of the horizon}
Subtract from the sextant altitude. $\text{Dip}=1.76\sqrt{h}\pr$, $h$ in metres.
\begin{center}\input{tables/corr_dip.tex}\end{center}

\section{Atmospheric refraction}
Subtract from the apparent altitude $H_a$. Bennett's formula; standard atmosphere.
\begin{center}\input{tables/corr_refr.tex}\end{center}

\section{Sun semidiameter}
Add for a lower-limb sight, subtract for an upper-limb sight.
\begin{center}\input{tables/corr_sd.tex}\end{center}

\chapter{GHA increments}

To find the GHA between whole-hour entries, add the increment for the elapsed
minutes. The Sun's GHA increases at $15\dg$ per hour; Aries at $15.041\dg$ per
hour. For the seconds, add $0.25\pr$ per second for the Sun and $0.2507\pr$ per
second for Aries.
\begin{center}\input{tables/corr_incr.tex}\end{center}

\backmatter
\chapter{Provenance and verification}
\sloppy
Every figure in this book was generated by
\texttt{almanac/}\allowbreak\texttt{generate\_tables.wls}
from the package \texttt{wolfram/}\allowbreak\texttt{CelestialNavigation.wl} in the repository
\url{https://github.com/mthiel74/SextantNavigation}, whose 186-test suite checks
the ephemeris against JPL Horizons and Meeus. Independent sanity checks performed
at generation time confirm: the Sun's declination reaches $+23.44\dg$ at the June
solstice and $-23.44\dg$ at the December solstice; the Sun's GHA advances
$15.00\dg$ per hour and Aries $15.041\dg$ per hour; a bright star's SHA is stable
to under $1\pr$ across the year; and the three worked examples close on the
engine's own geometry to $0.0\pr$. Re-generate with
\texttt{wolframscript -file}\allowbreak\ \texttt{almanac/}\allowbreak\texttt{generate\_tables.wls}
and rebuild with
\texttt{latexmk -pdf}\allowbreak\ \texttt{almanac/}\allowbreak\texttt{SextantAlmanac2026.tex}.

\medskip
\noindent\textbf{Remember the disclaimer:} educational use only, approximately one
arcminute accuracy, not for sole-means marine navigation.

\end{document}
